### 2016-06-27 reasoning.tex
edits

\noindent The soundness of rely-guarantee judgments can be stated
thus\footnote{Full proofs are provided in the supplementary material.}:
\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to
Definition~\ref{def:rg-semantics}. 
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b) the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of the RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.

>>>>

\begin{theorem}[\bfseries Soundness] 
The rely-guarantee judgments defined by the rules in
Fig.~\ref{fig:rg-rules} are sound with respect to
the semantics of Definition~\ref{def:rg-semantics}.\footnote{Full proofs are provided in the supplementary material.}
\end{theorem}
\noindent In particular, if $\rg{\I,P,R}{c}{G,Q}$ can be derived using
the rules of Fig.~\ref{fig:rg-rules}, then (a) every interleaved
multi-step reduction of $c$ starting from a trace that satisfies $P$
and $\I$, results in a trace that satisfies $Q$, and (b) the effect
that every small-step of $c$ has on the trace is contained in $G$.
Soundness of the RG judgment for transaction-bound commands
($\txnbox{c}_i$) is stated similarly.  For expressions, soundness of
the judgment $\rg{\I,P,R}{\txnbox{e}_i}{G,C,Q}$ also proves that $e$
is always evaluated to a value that satisfies $C$.

### 2016-06-27 store-consistency.tex
edits

The abstract machine of Fig.~\ref{fig:txnimp} allows operations to
witness arbitrary subsets of the global state, thus assuming the
semantics of an eventually consistent ({\sc ec}) data
store\footnote{{\sc ec} guarantees that in the absence of further
updates, all reads witness the same global state \emph{eventually}. In
any finite trace, however, there are no guarantees on what a read
witnesses.}. There are however stores, such as relational databases,
that provide stronger consistency guarantees than {\sc ec}. Like the
isolation levels of transactions, the consistency level of the
underlying store also affects the semantics of a program in
non-trivial ways. In this section, we demonstrate how the semantics of
stronger stores with on-demand weak isolation can be captured in our
operational model so that we can use the reasoning framework developed
in \S\ref{sec:reasoning} to reason about such stores.

@@

Firstly, we observe that the semantics of stronger stores can be
captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we split $\I$ into two components: (1).  $\I_s$,
the store-specific invariant, and (2). $\I_c$, the program-specific
(or, client-specific) invariant, to capture consistency and isolation
constraints, respectively. $\I$ is now a conjunction:
  $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often simplified to
  $\I \,=\, \I_s\wedge\I_c$).
While the program-specific trace invariant ($\I_c$) and remains an
invariant regardless of the store, the store specific invariant
($\I_s$) changes from store to store depending on the consisteny
level. $\I_s$ for an {\sc ec} store is $true$. Stronger stores, such
as an {\sc sc} store, have non-trivial $\I_s$.

A strongly-consistent {\sc sc} store guarantees a total order of all
operations w.r.t $\visZ$ consistent with their chronological order. A
straightforward $\I_s$ for this store is the \C{SC} property
formalized below:
\begin{smathpar}
\begin{array}{l}
  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
transactions execute under \iso{Read Committed} isolation. Imagine a
sample execution where $\eta_1$'s transaction is not yet committed
when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
The only way to satisfy both the invariants is to rule out all the
executions that interleave the operations of one transaction with the
other, thereby enforcing serializability \footnote{Thus,
serializability is a natural generalization of {\sc sc} to
transactions.}. In general, when $\I_s$ is in conflict (but not
inconsistent) with $\I_c$, the only way to enforce both invariant sets
is to restrict concurrency. Clearly, this is unacceptable since it
defeats the very purpose of supporting weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailisvldb} is often used in practice, perhaps to
capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
visiblity principle} to determine an acceptable weakening of $\I_s$ in
case of a conflict with $\I_c$.  The principle requires the weakened
consistency guarantee ($\I_s'$) of the store to enforce all visibility
relationships imposed by the actual consistency guarantee ($\I_s$),
unless enforcing such a relationship violates $\I_c$. 
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes relational database-like stores.

Semantics of a causally consistent data
store~\cite{gotsmanpopl16,LBC16} can also be obtained: by defining
$\I_s$ to be the causal consistency ({\sc cc}) invariant that is
appropriately weakened to resolve conflicts with $\I_c$. Details are
relegated to the supplementary in the interest of space.

\begin{remark}
Note that the purpose of maximum visibility principle is to
rationalize the observed behaviour of databases in practice. In
particular, it is not intended to be a guiding principle to engineer
data stores.
\end{remark}

@@

Reasoning under weakly consistent replication has received special
attention in the recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case. In this section, we explain how the standard artifacts
of weakly consistent replication manifest in our reasoning framework.

Reasoning about weakly consistent replication often involves reasoning
in terms of \emph{replicas} of data. The primary challenge in this
setting is to ensure that the assumptions made and guarantees enforced
by an operation at one replica carry over to other replicas that merge
its effects, thus preserving the overall integrity of the system.
In~\cite{gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the reasoning approach
from being applied to weaker stores such as Bayou~\cite{bayou} or
Quelea~\cite{pldi15}. Our view of replication however does not
explicitly involve replicas.  Fig.~\ref{fig:ec-theirs-vs-ours}
contrasts our model of weakly consistent replication with the
conventional replica-based model.  Under our model, the notion of a
replica is subsumed by the concept of visibility; a replica is defined
by the subset ($S$) of global state ($\E.\A$) that an operation
witnesses. Constraints over replica states therefore manifest as
constraints over the visibility relation.  For example, instead of
requiring the store to be causally consistent, an operation can
request to witness a causally consistent subset of the state; such
demands can be made via the trace invariant $\I$. For a precondition
($P$) of the operation to be useful, it has to be an assertion over
every causally consistent subset of the global state.  Since any
replica that eventually executes the operation has to expose one such
subset ($S$), the precondition is guaranteed to hold regardless of the
replica. There is however one problem with this explanation; by
considering subsets of just one global state, it ignores the fact that
the global state (hence, the replica states) change during the
execution of the operation. To account for this
change,~\cite{gotsmanpopl16} distinguishes between \emph{effect
generation} event at one replica and \emph{effect merge} event at
another replica, and requires that certain invariants be preserved
between these two events at different replicas. Our framework folds
all of this into a stability condition (\S\ref{sec:rely-guarantee}).
Since any change to the global state during the execution of the
operation is an interference, and $P$ is required to be stable with
respect to any such interference, it follows that $P$ is valid on
every replica, thus ensuring that assumptions made at the generation
event are also valid at the merge event.

@@

The second part of the proof is to show that assertions are stable
despite the interference from the the concurrent thread executing
\C{Wd2}. The interference is given by the following rely relation
($R_1$):
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}
The rely relation says the following about the concurrent thread: (1).
It may interfere only if \C{Wd2} is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits \C{Wd2}, then \C{Wd2} should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not \C{Wd1} has already committed.
Stability for most assertions is straightforward. The only non-trivial
proof is for $\phi_5$, where {\sc si} condition should be used to
show that any interference from \C{Wd2} is invalid. The proof proceeds
on the same lines as the Hoare proof discussed above; it is also
discussed in \S\ref{sec:motivation}.

>>>>

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicing the
behaviour of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.

@@

First, we observe that the semantics of a weakly-consistent data store
can be captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, resp.  $\I$ is now a
conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
trace invariant ($\I_c$) remains an invariant regardless of the
specific consistency and visibility features exposed by the underlying
data store, the store specific invariant ($\I_s$) changes from store
to store depending on the consisteny level.  For an eventually
consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
as those that support strong consistenty, have non-trivial definitions
for $\I_s$.

A strongly-consistent {\sc sc} store guarantees a total order of all
operations w.r.t $\visZ$ consistent with their chronological order. A
straightforward $\I_s$ for this store is the \C{SC} property
formalized below:
\begin{smathpar}
\begin{array}{l}
  \C{sc}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
transactions execute under \iso{Read Committed} isolation. Imagine a
sample execution where $\eta_1$'s transaction is not yet committed
when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
The only way to satisfy both invariants is to rule out all 
executions that interleave the operations of one transaction with the
other, thereby enforcing serializability.  In general, when $\I_s$ conflicts (but is not
inconsistent) with $\I_c$, the only way to enforce both invariant sets
is to restrict concurrency. Clearly, this is unacceptable since it
defeats the very purpose of supporting weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailisvldb} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visiblity principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes many kinds of relational database-like stores.

For example, the semantics of a causally consistent data
store~\cite{gotsmanpopl16,LBC16} can also be obtained by defining
$\I_s$ to be the causal consistency ({\sc cc}) invariant that is
appropriately weakened to resolve conflicts with $\I_c$.\footnote{Details
of this instantiation are relegated to the supplementary in the interest of space.}

%% \begin{remark}
%% Note that the purpose of maximum visibility principle is to
%% rationalize the observed behaviour of databases in practice. In
%% particular, it is not intended to be a guiding principle to engineer
%% data stores.
%% \end{remark}

@@

Reasoning under weakly consistent replication has received special
attention in recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case. In this section, we explain how the standard artifacts
of weakly-consistent replication manifest in our reasoning framework.

Reasoning about weakly-consistent replication involves reasoning in
terms of \emph{replicas} of data. The primary challenge in this
setting is to ensure that the assumptions made and guarantees enforced
by an operation at one replica carry over to other replicas that merge
their effects, thus preserving the overall integrity of the system.
In~\cite{gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the reasoning approach
from being applied to stores that provide guarantees weaker than
causal consistency (e.g., \cite{bayou,pldi15}).

Notably, our view of replication however does not explicitly involve
replicas.  Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based
model.  Under our model, the notion of a replica is subsumed by the
concept of visibility; a replica is defined by the subset ($S$) of
global state ($\E.\A$) that an operation witnesses. Constraints over
replica states therefore manifest as constraints over a specific visibility
relation.  For example, instead of requiring the store to be causally
consistent, an operation can request to witness a causally consistent
subset of the state; such demands can be made via the trace invariant
$\I$. For a precondition ($P$) of the operation to be useful, it has
to be an assertion over every causally consistent subset of the global
state.  Since any replica that eventually executes the operation has
to expose one such subset ($S$), the precondition is guaranteed to
hold regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for this
change,~\cite{gotsmanpopl16} distinguishes between an \emph{effect
  generation} event at one replica and an \emph{effect merge} event at
another replica, and requires that certain invariants be preserved
between these two events at different replicas. Our framework folds
all such machinery into a stability condition (\S\ref{sec:rely-guarantee}).
Since any change to the global state during the execution of the
operation is an interference, and $P$ is required to be stable with
respect to any such interference, it follows that $P$ is valid on
every replica, thus ensuring that assumptions made at the generation
event are also valid at the merge event.

@@

The second part of the proof is to show that assertions are stable
despite the interference from the the concurrent thread executing
\C{Wd2}. The interference is given by the following rely relation
($R_1$):

\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}

\noindent The rely relation says the following about the concurrent thread: (1).
It may interfere only if \C{Wd2} is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits \C{Wd2}, then \C{Wd2} should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not \C{Wd1} has already committed.
Stability for most assertions is straightforward. The only non-trivial
proof is for $\phi_5$, where {\sc si} condition should be used to
show that any interference from \C{Wd2} is invalid. The proof proceeds
on the same lines as the Hoare proof discussed above; it is also
discussed in \S\ref{sec:motivation}.
